Follow-up Samples

### 2017-02-16 submission/manuscript_outline_20161024.Rmd
version for lab review

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currenlty the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].

Recently, there has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have also shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC. Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgerical removal of lesion (adenoma or carcinoma) and whether the community changes at all.

In this study we tested the hypothesis that the bacterial microbiome changes after surgery for individuals with a lesion.  Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed pre (initial) and post (follow up) surgery.  We also used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.

@@

***Positive Probability after Lesion Removal:***
If there were common OTUs for individuals with adenoma and carcinoma, that were different versus normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or an initial sample. This is what we observed regardless of model used (lesion or initial sample) or whether it was built on the full or reduced variable data set [Figure 4 & S4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).

When we seperated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05). We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).  This lack of difference between the predicted and actual classifications were also observed for the initial sample model (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05). Even though the lesion model was not as accurate in classifying samples as the initial sample model. It was able to correctly keep the one individual who still had a carcinoma on follow up above the cut off threshold for a positive call [Figure 4a & S4a] while the initial sample model did not [Figure 4b & S4b].

@@

## Discussion
From our results there were some large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There was much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].

@@

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial samples in order to find common OTUs in the community that change for both adenoma and carcinoma.  What we found was that the commonly associated CRC bacteria were not higly represented within our models but rather that OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and before sample) significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that it is possible that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

Many of the common OTUs between the different models used had many OTUs that taxonomically classifed to potential butyrate producers [Table S2].  Another batch of OTUs classifed to bacteria that can either degrade polyphenols or are inhibited by them.  Both butyrate and polyphenols are thought to protective against cancer in part by reducing inflammation [@louis_gut_2014].   These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.  One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the exisiting community but also have an effect on CRC risk and disease progression.

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our lesion model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.  Another limitation is that this study drew heavily from those with Caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be immediately extrapolate-able to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.  The time difference in collection of sample between adenoma and carcinoma could have affected our observed results for the differences between individuals with adenoma or carcinoma.  This confounding though does not affect the observations based on the overall lesion results.

Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occurred because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.

Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after surgical removal of lesion.  Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.  Our data provides evidence that commensal bacteria are important in the development of polyps and also potentially the transition from adenoma to carcinoma.

@@

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to normalize within this time period [@obrien_impact_2013].  Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***FIT and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013]. 16S rRNA gene sequencing was completed as previously described by Kozich, et al. [@kozich_development_2013].  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described [@kozich_development_2013].  The general processing workflow using mothur was as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

***Lesion Model Creation***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  The model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyper-parameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.  Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.

***Before Sample Model Creation***
We also investigated whether a model could be created that could identify before and after surgery samples.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure that was used to create the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being a before surgery sample.

***Selection of Important OTUs***
In order to assess which variables were most important to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion or before sample models.

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.

***Analysis Overview***
Initial and follow up samples were analyzed for differences in alpha and beta diversity.  Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.  From here, all OTUs that were used in either model were then analyzed using a paired Wilcoxson test.  We then investigated the relative abundance of specific previously associated CRC bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.  We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the before sample model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature before sample model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, a list of common OTUs were found for the two different models used.

***Reproducible Methods.***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.  Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.

@@

**Figure 1: General Differences between the Adenoma or Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  B) A significant difference was found between the adenoma and carcinoma group for change in FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`). C) NMDS of the initial and follow up samples for the Adenoma group.  D) NMDS of the initial and follow up samples for the Carcinoma group. For C) and D) the teal represents initial samples and the pink represents follow up samples.

**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples. There was a significant difference in initial and follow up sample for the OTUs classified as *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).  B) Adenoma initial and follow up samples.  There were no significant differences between initial and follow up.

**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Before Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.  The blue line represents the reduced variable model using 100% of the data set.  A) Lesion model.  B) Before sample model

**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Before Sample Reduced Variable Models**  A) Lesion positive probability adjustment of those with carcinoma from initial to follow up sample  B) Initial follow up positive probability adjustment of those with carcinoma from initial to follow up sample  C) Lesion positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.    D) Initial follow up positive probability adjustment of those with adenoma as well as those with SRN and the probability adjustment from initial to follow up sample.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

@@

**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of OTUs in Initial versus Follow Up**

@@

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.

>>>>

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].

Recently, there has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have also shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC. Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgical removal of lesion (adenoma or carcinoma) and whether the community changes at all.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples after surgery for individuals with a lesion (adenoma or carcinoma).  Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest models and observed how these models, as well as specific OTUs within them, performed pre and post surgery.  We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.

@@

***Positive Probability after Lesion Removal:***
If there were common OTUs that could separate adenoma and carcinoma from normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or an initial sample. This is what we observed regardless of model used (lesion or initial sample) or whether it was built on the full or reduced variable data set [Figure 4 & S4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).

When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05). We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).  This lack of difference between the predicted and actual classifications were also observed for the initial sample model (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05). Even though the lesion model was not as accurate in classifying samples as the initial sample model. It was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 4a & S4a] while the initial sample model did not [Figure 4b & S4b].

@@

## Discussion
From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].  Although there was a detectable change in the carcinoma group.  It should be noted that whether this change was towards what wouldbe be expected for normal controls is not known.

@@

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial samples in order to find common OTUs in the community that change for both adenoma and carcinoma.  What we found was that the commonly associated CRC bacteria were not highly represented within our models but rather that OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT, both models (lesion and initial sample) significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community [Table S3].

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

Many of the common OTUs between the different models used had many OTUs that taxonomically classified to potential butyrate producers [Table S2].  Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them.  Both butyrate and polyphenols are thought to be protective against cancer in part by reducing inflammation [@louis_gut_2014].   These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes.  One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our lesion model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.  This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.  Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.  The time difference in collection of sample between adenoma and carcinoma could have affected our observations for differences between individuals with adenoma or carcinoma.  This confounding though would not affect the observations where these individuals are grouped together.

Interestingly, within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model alone.  Further independent studies need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.

Despite these limitations our findings add to the existing scientific knowledge on CRC and the bacterial microbiome: That there is a measurable difference in the bacterial community after surgical removal of lesion.  Further, the ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.  Our data provides evidence that commensal bacteria may be important in the development of polyps and also potentially the transition from adenoma to carcinoma.

@@

***Study Design and Patient Sampling:***
The sampling and design were similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to normalize within this time period [@obrien_impact_2013].  Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***FIT and 16S rRNA Gene Sequencing:***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013]. 16S rRNA gene sequencing was completed as previously described by Kozich, et al. [@kozich_development_2013].  DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

***Sequence Processing:***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described [@kozich_development_2013].  The general processing workflow using mothur was as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

***Lesion Model Creation:***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  In brief, the model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyper-parameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.  Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI)) - length(rownames(good_metaf))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.

***Initial Sample Model Creation:***
We also investigated whether a model could be created that could identify pre (initial) and post (follow up) surgery samples.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial surgery sample.

***Selection of Important OTUs:***
In order to assess which variables were most important to either model we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different 80/20 split models and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion or before sample models.

***Statistical Analysis:***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.  Where multiple comparison testing was appropriate a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs that were used in either built model, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the initial sample model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial sample model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  A list of common OTUs were found for the two different models used.  Finally, both lesion and initial sample models initial positive probabilites were tested against the follow up positive probabilities.

***Reproducible Methods:***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.  Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.

@@

**Figure 1: General Differences between the Adenoma or Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  B) A significant difference was found between the adenoma and carcinoma group for change in FIT measurement (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`). C) NMDS of the initial and follow up samples for the Adenoma group.  D) NMDS of the initial and follow up samples for the Carcinoma group. For C) and D) the teal represents initial samples and the pink represents follow up samples.

**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples had an observed significant difference in initial and follow up sample for the OTUs classified as *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).  B) Adenoma initial and follow up samples.  There were no significant differences between initial and follow up.

**Figure 3: Receiver Operating Characteristic Curve for Lesion and Initial Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.  The blue line represents the reduced variable model using 100% of the data set and what was used for subsequent classification.  A) Lesion model.  B) Before sample model

**Figure 4: Breakdown by Carcinoma and Adenoma of Prediction Results for Lesion and Initial Sample Reduced Variable Models**  A) Lesion positive probability change from initial to follow up sample in those with carcinoma   B) Initial positive probability change from initial to follow up sample in those with carcinoma   C) Lesion positive probability change from initial to follow up sample if those with adenoma or Screen Relevent Neoplasia (SRN).    D) Initial positive probability change from initial to follow up sample in those with adenoma or SRN.  The dotted line represents the threshold used to make the decision of whether a sample was positive or not.

@@

**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of All OTUs for Initial versus Follow Up**

@@

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Amanda Elmore for reviewing and correcting code error and providing feedback on manuscript drafts.  We would also like to thank Nicholas Lesniak for providing feedback on manuscript drafts.
